Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-07-08 MS/ms.Rmd
More writing. Fixing heterogenity section, introducing some important ideas on PIs

<!--- As said before you want to play up the aspect of investigaing heterogeneity rather than controlling for it - controlling for it seems to mean you still want to know the overall effect after getting rid of every confunders. Whilst it is also what we want, we really want to explain the hetero --->

It will not be possible to derive an effect size that completely controls for known heterogeneity at the effect size level all the time. This limitation is partly because treatment effects are not be simple linear or exponential functions of the mean response making the effect sizes we discuss above unsuitable [e.g., thermal performance curves; @Noble2018]. For example, dosages applied to treatments can follow quite complex non-linear relationships, with exponential or quadratic relationships in relation to some mean response. When this occurs we recommend comparative physiologists apply meta-regression approaches, controlling for these known treatment differences as a moderator. There are simple statistical approaches a meta-analyst can use to then make the overall effects interpretable [@Schielzeth2010]. Before diving into meta-regression models, we will first overview the meta-analytic approaches or steps commonly used.

@@

Assume that we are interested in understanding the overall effect of, say, stress (i.e., corticosterone - CORT) on reproduction [e.g., see @Podmola2018] we might extract data from studies manipulating corticosterone in bird eggs. Of course, these studies are diverse, apply different dosages of corticosterone and are working across many species. If the studies are experimental in nature, than a standardised effect size, such as $ln_{RR}$ or Hedges' *d* might be applicable. We then might fit an overall, multilevel meta-analytic model as follows: 
$$
\begin{equation} 
Y_{i} = \beta_{0} + a_{k[i]} + sp_{k[i]} + s_{j[i]} + e_{i} + m_{i}\
(\#eq:MLMM)
\end{equation} 
$$ 
$$
\begin{equation} 
a_{k[i]}\sim N(0, \sigma^{2}_aA),\
sp_{k[i]}\sim N(0, \sigma^{2}_{sp}I),\
s_{j[i]}\sim N(0, \sigma^{2}_{sp}I),\
e_{i}\sim N(0, \sigma^{2}_eI),\
m_{i} \sim N(0, v_{i})\
(\#eq:dist)
\end{equation} 
$$

@@

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i*. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_aA$ [i.e., *N*$(0, \sigma^{2}_aA)$], where A is a phylogenetic correlation matrix derived from a phylogenetic tree, $sp_{k[i]}$ is the species-specific effect for the *k*th species applied to effect size *i*, $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect sizes, $v_{i}$, which is known. This model is a 'weighted' multilevel meta-analysis model because the estimation of *u* is weighted by the inverse sampling variance of each effect size. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation such as differences in the methods applied, species differences or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our overall question: Across all the species and studies manipulating egg corticosterone, how much of an overall effect does corticosterone have on reproduction relative to control conditions? <!--- don't we need to mention in the case of VCV for mi--->

@@

Quantifying effect size heterogeneity (i.e., effect size variance) provides a pathway to explore hypotheses thought to explain why effects might vary within and across studies. Effect size heterogeneity can be thought of as the percentage of variance between effect sizes that cannot be attributable to sampling error (often as called 'total' heterogeneity) [@Higgins2002]. More formally, we can quantify heterogeneity using the multilevel meta-analytic model we describe above as:

@@

where $\sigma^2_{total} = \sigma^2_{study} + \sigma^2_{phylogeny} + \sigma^2_{species} + \sigma^2_{residual} +\sigma^2_{m}$ is the total effect size variance (using the above multilevel model) and $\sigma^2_{m}$ is 'typical' the sampling error variance. <!--- you need more than that sigma_m is a typical one - you may need to provide a formula here --->. Quantifying heterogeneity provides a formal way to assess how much variation is the result of 'true' biological or methodological differences as opposed to simply chance (i.e., sampling variance) [@NakagawaSantos2012; @Higgins2002]. We can only conduct such analyses if we have an estimate of the sampling variance for a given effect measure because only then can we quantify how much variation is the result of sampling variability. @Senior2016 have shown that total heterogeneity is often extremely high in ecological and evolutionary meta-analysis (\~91%), suggesting that effect measures vary widely across and within studies. Some of the variation is clearly the result of working with different species or strains, different traits and different outcome measures, however, some of this variation also arises because of methodological differences across studies that might be related to the dosages or temperatures used for treatments. Formally quantifying and presenting heterogeneity estimates from meta-analytic models can provide a way to understand the major drivers of effect size variation by allowing researchers to produce various $I^2$ estimates which can be formally compared within and across studies. For example, we could fit the multilevel model described above to understand what proportion of variation is the result of, say, 'study' effects following @NakagawaSantos2012: $I^2_{study} = \frac{\sigma^2_{study}}{\sigma^2_{total}}$ where $\sigma^2_{study}$ is the study specific variance, and all other variance components are for phylogeny, species, residual (effect size level) and m (total measurement/sampling error). Many readers will notice the similarities between $I^2$ and $R^2$ [@Nakgawa2013; @Nakagawa2010]. If we want to formally assess how much effect size variation is the result of methodological factors, such as temperature and dosage differences applied across studies, then we could fit our multilevel meta-regression model, including dosage or temperature (linear, quadratic or even more complex fits), and estimate how much variation is explained by these fixed effects following @Nakgawa2013.

@@

Note that this formula does not include $\sigma^2_{m}$ as sampling error variance is assumed to be known in meta-analysis as explained above <!--- if you have not done so, please do --->. The $R^2_{marginal}$ provides a formal means to assess just how much variation in effects is the result of known heterogeneity at the effect size level <!--- note what level's variation it accounts for depends on the type of a fixed effect or moderator --->, driven by continuous moderators such as dosage, temperature, salinity, and even pH. $R^2_{marginal}$, or how we like to think of it, $R^2_{heterogenity}$, maybe a useful measure to provide to help readers understand just how much effect variation can result from the specific treatment application. If the heterogeneity is high as a result, say, dosages differences, then it is clear that the choice of dosage within and across studies will have a critical impact on the effect of interest. <!--- you can also say metaAider and orcharRd packages have functions to cacualte this already via metafor - implemention is important --->

>>>>

It will not be possible to derive an effect size that completely controls for 'nuisance' heterogeneity at the effect size level all the time. This limitation is partly because treatment effects may not be simple linear or exponential functions of the mean response making the effect sizes we discuss above unsuitable [e.g., thermal performance curves; @Noble2018]. For example, dosages applied to treatments can follow quite complex non-linear relationships (e.g., quadratic relationships) in relation to some mean response. When this occurs we recommend comparative physiologists apply meta-regression approaches, controlling for 'these known treatment differences 'nusiance heterogenity' as a moderator. In many cases, this will be the easiest, or even preferred approach because exploring drivers of heterogenity will be the main interest in most meta-analyses. Nonetheless, there are some simple statistical approaches a meta-analyst can use to then make the overall effects interpretable, allowing one to control for 'nuisance heterogenity', while also allowing other biological moderators of interest to be tested [@Schielzeth2010]. Before diving into meta-regression models, we will first overview the meta-analytic approaches or steps commonly used.

@@

Assume that we are interested in understanding the overall effect of, say, stress (i.e., corticosterone - CORT) on reproduction [e.g., see @Podmola2018] we might extract data from studies manipulating corticosterone in bird eggs. Of course, these studies are diverse, apply different dosages of corticosterone and are working across many species. If the studies are experimental in nature, than a standardised effect size, such as $ln_{RR}$ or Hedges' *d* might be applicable. We then might fit an overall, multilevel meta-analytic model as follows: 
$$
\begin{equation} 
Y_{i} = \beta_{0} + a_{k[i]} + sp_{k[i]} + s_{j[i]} + e_{i} + m_{i}\
(\#eq:MLMM)
\end{equation} 
$$ 
$$
\begin{equation} 
a_{k[i]}\sim N(0, \sigma^{2}_{phylogeny}A),\
sp_{k[i]}\sim N(0, \sigma^{2}_{species}I),\
s_{j[i]}\sim N(0, \sigma^{2}_{study}I),\
e_{i}\sim N(0, \sigma^{2}_{residual}I),\
m_{i} \sim N(0, v_{i})\
(\#eq:dist)
\end{equation} 
$$

@@

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i*. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}A)$], where A is a phylogenetic correlation matrix derived from a phylogenetic tree, $sp_{k[i]}$ is the species-specific effect for the *k*th species applied to effect size *i*, $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect sizes, $v_{i}$, which is known as the sampling variance for effect. This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is weighted by the inverse sampling variance of each effect size. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation such as differences in the methods applied, species differences or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our overall question: Across all the species and studies manipulating egg corticosterone, how much of an overall effect does corticosterone have on reproduction relative to control conditions? <!--- don't we need to mention in the case of VCV for mi--->

@@

Quantifying effect size heterogeneity (i.e., effect size variance) provides the impetus for exploring hypotheses that are thought to explain why effects might vary within and across studies, while also providing context for interpreting the generality of effects. There are a number of measures commonly used to quantify effect size heterogenity [@Borenstein2019], however, in ecological and evolutionary meta-analysis, heterogeneity is often calculated and reported as $I_{total}^2$ as follows (if using a multilevel meta-analytic model as we describe above):

@@

where $\sigma^2_{total} = \sigma^2_{study} + \sigma^2_{phylogeny} + \sigma^2_{species} + \sigma^2_{residual} +\sigma^2_{m}$ is the total effect size variance  and $\sigma^2_{m}$ is the sampling error variance calculated as:
$$

$$
Quantifying heterogeneity using $I_{total}^2$ provides a formal way to assess the relative amount of variation that is the result of 'true' biological or methodological differences as opposed to simply chance (i.e., sampling variance) [@NakagawaSantos2012; @Higgins2002]. In other words, $I_{total}^2$ can be thought of as the percentage of variance between effect sizes that cannot be attributable to sampling error [@Higgins2002]. We can only calculate this metric if we have an estimate of the sampling variance for a given effect measure because only then can we quantify the proportion of variation that is the result of sampling variability. @Senior2016 have shown that total heterogeneity is often extremely high in ecological and evolutionary meta-analysis (\~91%), suggesting that effect measures vary widely across and within studies. Some of the variation is clearly the result of working with different species or strains, different traits and different outcome measures, however, some of this variation also arises because of methodological differences across studies that might be related to the dosages or temperatures used for treatments. Formally quantifying and presenting heterogeneity estimates from meta-analytic models can provide a way to understand the major drivers of effect size variation by allowing researchers to produce various $I^2$ estimates which can be formally compared within and across studies. For example, we could fit the multilevel model described above to understand what proportion of variation is the result of, say, 'study' effects following @NakagawaSantos2012: $I^2_{study} = \frac{\sigma^2_{study}}{\sigma^2_{total}}$ where $\sigma^2_{study}$ is the study specific variance, and all other variance components are for phylogeny, species, residual (effect size level) and m (total measurement/sampling error). Many readers will notice the similarities between $I^2$ and $R^2$ [@Nakgawa2013; @Nakagawa2010]. If we want to formally assess how much effect size variation is the result of methodological factors, such as temperature and dosage differences applied across studies, then we could fit our multilevel meta-regression model, including dosage or temperature (linear, quadratic or even more complex fits), and estimate how much variation is explained by these fixed effects following @Nakgawa2013.

@@

Note that this formula does not include $\sigma^2_{m}$ as sampling error variance is assumed to be known in meta-analysis as explained above. When including continuous moderators, such as dosage, temperature, salinity, and even pH, $R^2_{marginal}$ provides a formal means to assess just how much variation in effects is the result of 'nuisance' heterogeneity at the effect size level. Other biological moderators of interest could also be included as fixed effects and $R^2_{marginal}$ calculated for each independently or all together. Importantly, moderators could be those that explain variation at the effect size level (i.e., dosage, temperature) or they could explain variation at the study or species level (e.g., reproductive mode, endothermy etc). $R^2_{marginal}$, or how we like to think of it, $R^2_{'nuisance'}$, maybe a useful measure to help readers understand just how much effect variation can result from the specific treatment application. If the heterogeneity is high as a result of, say, dosages differences, then it is clear that the choice of dosage within and across studies will have a critical impact on the effect of interest. Comparative physiologists intersted in implementing thse calculations can do so using our packages `orchaRd` [@Nakagawa2021c] or `metaAidR` (https://github.com/daniel1noble/metaAidR).

While heterogenity estimates using $I^2$ measures can be useful for understanding the relative contributions of different factors to effect size variation, prediction intervals might be more appropriate in many cases [@Nakagawa2021c; @Borenstein2019]. Unlike $I^2$, prediction intervals are more holistic and probably more meaningful in the context of meta-analysis. Prediction intervals make use of the standard deviation of the mean instead of the standard error (which creates confidence intervals) [@Borenstein2019].
